---
title: "Final Project Code"
author: "Kelsey Matsik"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning

```{r}
library(dplyr)

d1 <- read.csv("~/Desktop/STAT 3280/Project/Data/student-mat.csv", header=TRUE)
d2 <- read.csv("~/Desktop/STAT 3280/Project/Data/student-por.csv", header=TRUE)

print(colnames(d1))
print(colnames(d2))

# View dimensions of the dataframe 
print(dim(d1))
print(dim(d2))

# Combine d1 and d2
combined_df <- rbind(d1, d2)

# Check for duplicate rows 
duplicate_rows <- combined_df[duplicated(combined_df),]
duplicate_rows # No duplicates 

# Rename columns to be more intuitive 
df <- combined_df 

df <- df %>% rename(address.type = address, 
                    divorced = Pstatus, 
                    mom.edu = Medu, 
                    dad.edu = Fedu, 
                    mom.occ = Mjob, 
                    dad.occ = Fjob, 
                    higher.edu = higher, 
                    home.internet = internet, 
                    weekday.alc = Dalc, 
                    weekend.alc = Walc, 
                    health.status = health,
                    final.grade = G3, 
                    first.period.grade = G1, 
                    second.period.grade = G2
                    )

df1 <- df 

# Replacing variable values with original values
df1$famsize[df1$famsize == "LE3"] <- "less than or equal to 3"
df1$famsize[df1$famsize == "GT3"] <- "greater than 3"

df1$divorced[df1$divorced == "T"] <- 0
df1$divorced[df1$divorced == "A"] <- 1

df1$mom.edu[df1$mom.edu == 0] <- "no education"
df1$mom.edu[df1$mom.edu == 1] <- "elementary school"
df1$mom.edu[df1$mom.edu == 2] <- "middle school"
df1$mom.edu[df1$mom.edu == 3] <- "high school"
df1$mom.edu[df1$mom.edu == 4] <- "university or higher"

df1$dad.edu[df1$dad.edu == 0] <- "no education"
df1$dad.edu[df1$dad.edu == 1] <- "elementary school"
df1$dad.edu[df1$dad.edu == 2] <- "middle school"
df1$dad.edu[df1$dad.edu == 3] <- "high school"
df1$dad.edu[df1$dad.edu == 4] <- "university or higher"

# Changing 'yes/no' values to binary 1/0 values 
df1 <- df1 %>%
  mutate(across(c(schoolsup, famsup, paid, activities, nursery, higher.edu, 
                  home.internet, romantic), ~ ifelse(. == 'yes', 1, 0)))


# Transforming final.grade scale to 0-100 to be more comparable to real scenarios
df1 <- df1 %>% mutate(final.grade = final.grade * 5) # Values range from 0 to 20, so just multiply each one by 5
df1 <- df1 %>% mutate(first.period.grade = first.period.grade * 5)
df1 <- df1 %>% mutate(second.period.grade = second.period.grade *5)

# Making one drinking column 
df1$drinking1 <- df1$weekday.alc * 5 # Assigning weights for number of days in week and weekend
df1$drinking2 <- df1$weekend.alc * 2 

df1 <- df1 %>% mutate(drinking = rowMeans(select(., c("weekday.alc", "weekend.alc"))))

write.csv(df1, "~/Desktop/STAT 3280/Project/cleaned_data.csv", row.names=FALSE)
```

# Visualizations 

```{r}
df <- read.csv("~/Desktop/STAT 3280/Project/cleaned_data.csv")
```

```{r}
library(ggplot2)
library(viridis)
library(dplyr)
library(plotly)
library(knitr)
```


## Plot 1 

```{r}
library(ggpubr)

p9a <- ggplot(df, aes(x = age, y = drinking, color = drinking)) +
  geom_point(position = "jitter") + 
  scale_color_viridis(option = "plasma") + 
  theme_minimal() +
  geom_smooth(method = "lm", color = "black") +
  stat_cor(method = "pearson", r.accuracy = 0.01, p.accuracy = 0.001, label.x.npc = 0.8) + 
  # facet_grid(.~sex, labeller = as_labeller(c("F" = "Female", "M" = "Male"))) + 
  labs(title = "Age vs. Alcohol Consumption", 
       x = "Age", y = "Alcohol Consumption Level") + 
  scale_y_continuous(breaks = 1:5, labels = c("Very low", "Low", "Moderate", "High", "Very high")) + 
  theme(axis.text.y = element_text(), 
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 14), 
        strip.text = element_text(size = 14),
        plot.title = element_text(size = 18, hjust = 0.5)) + 
  guides(color = FALSE)

p9a
```

## Plot 2 

```{r}
library(ggpubr)

p9 <- ggplot(df, aes(x = age, y = drinking, color = drinking)) +
  geom_point(position = "jitter") + 
  scale_color_viridis(option = "plasma") + 
  theme_minimal() +
  geom_smooth(method = "lm", color = "black") +
  stat_cor(method = "pearson", r.accuracy = 0.01, p.accuracy = 0.001, label.x.npc = 0.5) + 
  facet_grid(.~sex, labeller = as_labeller(c("F" = "Female", "M" = "Male"))) + 
  labs(title = "Age vs. Alcohol Consumption", 
       x = "Age", y = "Alcohol Consumption Level") + 
  scale_y_continuous(breaks = 1:5, labels = c("Very low", "Low", "Moderate", "High", "Very high")) + 
  theme(axis.text.y = element_text(), 
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 14), 
        strip.text = element_text(size = 14),
        plot.title = element_text(size = 18, hjust = 0.5)) + 
  guides(color = FALSE)

p9
```

## Plot 3 
```{r}

library(shiny)
library(ggplot2)
library(viridis)

colors <- c("gold2", "orange2", "darkorange", "salmon3", 
            "orchid4", "purple3", "blue3", "blue4", "midnightblue")

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("binwidth", "Bin Width:", min = 0.5, max = 2, value = 1, step = 0.1),
      selectInput(inputId = "color",
                  label = "Fill Color",
                  choices = colors,
                  selected = "blue4"),  
      checkboxInput("facet_gender", "Group by Sex", value=FALSE) # Facet by gender
    ),
    mainPanel(
      plotOutput("plot", height="600px"), 
      uiOutput("annotation") # Output annotation when plot renders
    )
  )
)

server <- function(input, output) {
  df <- read.csv("cleaned_data.csv") 
  
  output$plot <- renderPlot({
    p <- ggplot(df, aes(x=drinking, y=health.status)) + 
      stat_bin_2d(binwidth = input$binwidth, color = "black") + # Change the bin width
      scale_fill_gradient(low = "white", high = input$color) +
      theme_minimal() + 
      labs(title = "Alcohol Consumption vs. Health Status", 
           x = "Alcohol Consumption Level", y = "Health Status", fill = "Count") + 
      scale_x_continuous(breaks = 1:5, labels = 
                           c("Very low", "Low", "Moderate", "High", "Very high")) +
      scale_y_continuous(breaks = 1:5, labels = 
                           c("Very poor", "Poor", "Moderate", "Good", "Very good")) +  
      theme(plot.title = element_text(hjust = 0.5, size=22), 
            axis.title = element_text(size = 22),
            axis.text.x = element_text(size = 14, angle = 45, vjust = 0.6),
            axis.text.y = element_text(size = 14), 
            plot.margin = margin(100, 100, 0, 100)) # Create margin where annotation will go
  
    if (input$facet_gender) { # Facet by gender 
      p <- p + facet_grid(. ~ sex, labeller = labeller(sex=c("F"="Female", "M"="Male"))) + 
        theme(strip.text = element_text(size = 20))
    }
  
    p
  })


output$annotation <- renderUI({ 
  tagList(
    div(style = "text-align: center; margin-top: -575px; margin-bottom: 20px; font-size: 14px;", 
        "This app visualizes the relationship between high school students' 
        drinking habits and health status. View all the students together, or 
        group by gender!")
   )
 })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

```

## Plot 4

```{r}
library(shiny)
library(ggplot2)
library(ggridges)
library(viridis)

# Define UI
ui <- fluidPage(
  # titlePanel("Weekday vs. Weekend Alcohol Consumption"),
  sidebarLayout(
    sidebarPanel(
      selectInput("variable", "View weekend alcohol consumption instead:",
                  choices = c("Weekday Alcohol Consumption" = "weekday.alc",
                              "Weekend Alcohol Consumption" = "weekend.alc"),
                  selected = "weekday.alc")
    ),
    mainPanel(
      plotOutput("density_plot")
    )
  )
)

# Define server logic
server <- function(input, output) {
  
  df <- read.csv("cleaned_data.csv") 
  
  output$density_plot <- renderPlot({
    # Check which variable is selected
    variable <- switch(input$variable,
                       "weekday.alc" = "Weekday Alcohol Consumption",
                       "weekend.alc" = "Weekend Alcohol Consumption")
    
    # Create the plot
    ggplot(df, aes(x = final.grade, y = factor(.data[[input$variable]]), 
                   fill = factor(.data[[input$variable]]))) + 
      geom_density_ridges() + 
      scale_fill_viridis_d(5, option = "plasma") +
      theme_minimal() + 
      scale_y_discrete(breaks = 1:5, labels = c("Very low", "Low", "Moderate", 
                                                "High", "Very high")) +
      labs(y = "Alcohol Consumption Level", x = "Final Grade", title =
           paste("Effect of", variable, "on Final Grade")) + 
      theme(legend.position = "none", 
            plot.title = element_text(hjust=0.5, size=22), 
            axis.title = element_text(size=20), 
            axis.text.x = element_text(size=16), 
            axis.text.y = element_text(size=16))  +
      scale_x_continuous(breaks = seq(0, 100, by = 10)) 
  })
}

# Run the application
shinyApp(ui = ui, server = server)
```


## Plot 5 

```{r}

study.df <- df %>%
  mutate(studytime = case_when(
    studytime == 1 ~ "<2 hours",
    studytime == 2 ~ "2-5 hours",
    studytime == 3 ~ "5-10 hours",
    studytime == 4 ~ ">10 hours",
    TRUE ~ as.character(studytime)
  ))

study.df$studytime <- factor(study.df$studytime, labels = c("<2 hours", 
                                                            "2-5 hours", 
                                                            "5-10 hours", 
                                                            ">10 hours"))

p2colors <- viridisLite::viridis(4, option = "plasma")

p2 <- plot_ly(study.df, x = ~absences, y = ~final.grade, color = ~studytime, colors = p2colors,
               text = ~paste(studytime, "spent studying per week"), 
               hoverinfo = "text", type="scatter", mode="markers", 
              marker = list(opacity = 0.7, size=8)) %>% 
  layout(title = list(text = "Total Absences vs. Final Grade", size = 20), 
         xaxis = list(title = "Absences", titlefont = list(size = 16), tickfont = list(size = 14)),
         yaxis = list(title = "Final Grade", titlefont = list(size = 16), tickfont = list(size = 14)),
         showlegend = TRUE, 
         legend = list(title = list(text = "Study Time per Week"))) 

p2
```


## Plot 6

```{r}
library(tidyverse)

# Pre-processing 

grades_df <- df %>% select(first.period.grade, second.period.grade, final.grade, health.status)
grades_df <- gather(grades_df, period, grade, 1:3)
grades_df$Period <- grades_df$period
grades_df$Period <- factor(grades_df$Period, levels = c("first.period.grade", 
                                                        "second.period.grade", "final.grade"))

grades_df <- grades_df %>%
  mutate(Period = case_when(
    Period == "first.period.grade" ~ "1st Semester",
    Period == "second.period.grade" ~ "2nd Semester",
    Period == "final.grade" ~ "Final",
    TRUE ~ as.character(Period)
  ))

grades_df <- grades_df %>%
  mutate(health.status = case_when(
    health.status == 1 ~ "Very Good",
    health.status == 2 ~ "Good",
    health.status == 3 ~ "Moderate",
    health.status == 4 ~ "Poor",
    health.status == 5 ~ "Very Poor",
    TRUE ~ as.character(health.status)
  ))

grades_df$health.status <- factor(grades_df$health.status, levels = c("Very Good", 
                                        "Good", "Moderate", "Poor", "Very Poor"))
```

```{r fig.align = 'center'}
p5colors <- c("blue4", "blue4", "grey", "grey", "grey")

p5 <- plot_ly(grades_df, x = ~factor(health.status), y = ~grade, type = "violin", frame = ~Period,
             color = ~factor(health.status), colors = p5colors) %>% 
  layout(
    title = list(text = "Grade Distribution by Student's Health Condition over Semester", size = 20),
            xaxis = list(title = "Health Status", titlefont = list(size = 16), 
                         tickfont = list(size = 14)),
            yaxis = list(title = "Grade", range = c(0, 100), 
                         titlefont = list(size = 16), 
                         tickfont = list(size = 14), dtick = 10),
            boxmode = "group", 
            showlegend = FALSE) %>%
  animation_opts(frame = 1000,  # Controls speed of transition; higher number = slower
                 redraw = TRUE, # Redraws the plot in each frame
                 easing = "linear") %>% # FIX THIS PART
  
  layout(
   updatemenus = list( 
     list(
       type = "buttons", 
       showactive = FALSE,
       x = 0, 
       y = -0.05, 
       buttons = list(
         
        list(
           label = "Pause", # Name of the label
           method = "animate", # What do to when the button is clicked
           args = list(NULL, list(frame = list(duration = 0,  # Stops immediately
                                               redraw = TRUE), 
                                  mode = "immediate"))
         )
       )
     )
   )
 ) 
    

p5
```


## Plot 7

```{r}
p7 <- ggplot(df, aes(x=factor(freetime), y=final.grade)) + 
  geom_boxplot(aes(fill=factor(freetime))) + 
  theme_minimal() + 
  scale_fill_viridis_d(option = "plasma") +
  facet_grid(.~activities, 
             labeller = as_labeller(c("0" = "No Extracurriculars", "1" = "Extracurriculars"))) + 
  guides(fill = FALSE) + 
  labs(title = "Final Grade Distribution by How Freetime is Spent", 
       x = "Amount of Freetime after School", y = "Final Grade") + 
  scale_x_discrete(breaks = 1:5, labels = c("A lot less", "Less", "Some", "More", "A lot more")) + 
  theme(axis.text.x = element_text(angle = 35), 
        plot.title = element_text(hjust = 0.5, size = 18), 
        axis.title = element_text(size = 14),
        strip.text = element_text(size = 14), 
        axis.text = element_text(size = 10))


p7
```


## Plot 8

```{r}
library(shiny)
library(ggplot2)
library(viridis)
library(dplyr)

# Define UI
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("variable", "View by father's education level instead:",
                  choices = c("Mother's Education Level" = "mom.edu",
                              "Father's Education Level" = "dad.edu"),
                  selected = "mom.edu")
    ),
    mainPanel(
      plotOutput("barPlot")
    )
  )
)


# Define server logic
server <- function(input, output) {
  
  df <- read.csv("cleaned_data.csv") # Load and prep data 
  
  edu_df <- df %>%
    mutate(higher.edu = case_when(
      higher.edu == 0 ~ "No", 
      higher.edu == 1 ~ "Yes", 
      TRUE ~ as.character(higher.edu)
    ))
  
  edu_df <- edu_df %>%
    mutate(mom.edu = case_when(
      mom.edu == "no education" ~ "No education", 
      mom.edu == "elementary school" ~ "Elementary school",
      mom.edu == "middle school" ~ "Middle school",
      mom.edu == "high school" ~ "High school", 
      mom.edu == "university or higher" ~ "University or higher", 
      TRUE ~ as.character(mom.edu)
    ))
  
  edu_df <- edu_df %>%
    mutate(dad.edu = case_when(
      dad.edu == "no education" ~ "No education", 
      dad.edu == "elementary school" ~ "Elementary school",
      dad.edu == "middle school" ~ "Middle school",
      dad.edu == "high school" ~ "High school", 
      dad.edu == "university or higher" ~ "University or higher", 
      TRUE ~ as.character(dad.edu)
    ))
  
  edu_df$mom.edu <- factor(edu_df$mom.edu, levels = c("No education", 
                                                      "Elementary school", 
                                                      "Middle school", 
                                                      "High school", 
                                                      "University or higher"))
  edu_df$dad.edu <- factor(edu_df$dad.edu, levels = c("No education", 
                                                      "Elementary school", 
                                                      "Middle school", 
                                                      "High school", 
                                                      "University or higher"))
  
  
  output$barPlot <- renderPlot({
    
    # Check which variable is selected
    variable <- switch(input$variable,
                       "mom.edu" = "Mother's Education Level",
                       "dad.edu" = "Father's Education Level")
    
    # Create the plot
    ggplot(edu_df, aes(y=.data[[input$variable]], fill=factor(higher.edu))) + 
      geom_bar(position = "dodge") + 
      geom_text(stat = "count", aes(label = after_stat(count)), 
                position = position_dodge(width = 0.9), vjust = 0.6, 
                hjust = -0.3, size = 5) + 
      theme_minimal() + 
      labs(title = paste("Interest in Attending College by", variable), 
           x = "Count", y = variable, fill = "College") + 
      theme(plot.title = element_text(hjust = 0.5, size = 22), 
            axis.title = element_text(size = 20), 
            axis.text = element_text(size = 16), 
            # axis.text.x = element_blank(),
            legend.title = element_text(size = 16), 
            legend.text = element_text(size = 14)) + 
      scale_fill_manual(values = c("grey", "#0D0887FF")) 
      # guides(fill = FALSE)
  })
}


# Run the application
shinyApp(ui = ui, server = server)
```


## Plot 9

```{r}
library(ggalluvial)

alluv_df <- df %>% 
  group_by(romantic, famrel, health.status) %>% 
  summarize(freq = n())

alluv_df$romantic <- factor(alluv_df$romantic, labels = c("No", "Yes"))
alluv_df$famrel <- factor(alluv_df$famrel, labels = c("Very poor", "Poor", 
                                                      "Decent", "Good", "Very Good"))
```


```{r, fig.width = 8, fig.align = 'center'}
p8 <- ggplot(alluv_df, 
       aes(y = freq, axis2 = factor(romantic), axis1 = factor(famrel))) + 
         geom_alluvium(aes(fill = factor(famrel)), width = 0, knot.pos = 1/4, reverse = FALSE) + 
  guides(fill = "none") +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), reverse = FALSE, size = 3.25) +
  scale_x_continuous(breaks = 1:2, labels = c("Quality of Family Relationships", "Romantic Partner")) + 
  labs(title = "Satisfaction with Family and Romantic Relationships", y = "Frequency") + 
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal() + 
  theme(plot.title = element_text(size = 16, hjust = 0.5), 
        axis.title.y = element_text(size = 14), 
        axis.text.x = element_text(size = 13), 
        axis.text.y = element_text(size = 10)) 

p8 
```


# Appendix 

The appendix contains additional plots I made that were not included in my final 
project. The plots below either served as a rough draft for plots upon which further 
iterations were made, or they taught me what did *not* work when trying to visualize 
variables in my data.  

## Additional Plots 


```{r}
library(dplyr)
library(ggcorrplot)
library(viridis)
library(wesanderson)
library(RColorBrewer)
library(ggbeeswarm)
library(patchwork)
library(ggridges)
library(plotly)
```


About the dataset: \newline 
<https://www.kaggle.com/datasets/uciml/student-alcohol-consumption> 
```{r}
df <- read.csv("~/Desktop/STAT 3280/Project/cleaned_data.csv")
colnames(df)
```
```{r}
grades <- df %>% select(first.period.grade, second.period.grade, final.grade)
head(grades, 20)
```


```{r}
ggplot(df, aes(x=weekend.alc)) + geom_bar()
```

```{r}
ggplot(df, aes(x=weekday.alc)) + geom_bar()
```

## Exploratory Data Analysis 

```{r}
# Define numeric variables (not including binary variables)
num_vars <- df %>% select(age, traveltime, studytime, failures, famrel, 
                          freetime, goout, weekday.alc, weekend.alc, 
                          health.status, absences, final.grade, drinking)
```

Correlation Matrix (Numeric Variables)
```{r}
cor_matrix <- cor(num_vars)
cor_df <- as.data.frame(cor_matrix)

p1 <- ggcorrplot(cor_df, type = "lower", lab = TRUE) + 
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  labs(x = "Variable 1", y = "Variable 2", title = "Correlation between Numeric Variables", fill = "Correlation") + theme(plot.title = element_text(hjust=0.5))

p1_color_scale <- scale_fill_gradient2(low = "red", mid = "white", high = "green3", midpoint = 0, limits = c(-1, 1), na.value = "white")

p1 <- p1 + p1_color_scale

p1
```
Source: <http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2>

```{r}
p2 <- ggplot(df, aes(x=factor(drinking), y=final.grade, fill=factor(drinking))) + geom_boxplot() + theme_minimal() + labs(title="Drinking Habits vs. Final Grade") +
scale_fill_manual(values=wes_palette("Darjeeling1", n=10, type="continuous"))
p2
```

```{r}
p3 <- ggplot(df,aes(x=factor(drinking), y=final.grade, fill=factor(drinking))) + 
  geom_violin() +
scale_fill_manual(values=wes_palette("Darjeeling1", n=10, type="continuous")) + 
  labs(title="Drinking Habits vs. Final Grade")
p3
```

```{r}
p4 <- ggplot(df, aes(x=final.grade, fill=mom.edu)) + 
  geom_density(alpha=0.5) + 
  scale_fill_manual(values=wes_palette("Darjeeling1", n=10, type="continuous")) + 
  theme_minimal() + 
  labs(title="Final Grade Distribution grouped by Mother's Education")

p4
```

```{r}
p5 <- ggplot(df, aes(x=final.grade, fill=dad.edu)) + 
  geom_density(alpha=0.5) + 
  scale_fill_manual(values=wes_palette("Darjeeling1", n=10, type="continuous")) + 
  theme_minimal() + 
  labs(title="Final Grade Distribution grouped by Fathers's Education")

p5
```

```{r}
p6 <- p4 + p5
p6
```

```{r}
p7 <- ggplot(df, aes(x=divorced, y=final.grade, fill=factor(divorced))) + geom_boxplot()
p7
```


```{r}
p8 <- ggplot(df, aes(x=schoolsup, y=final.grade, fill=factor(divorced))) + geom_boxplot()
p8
```

```{r}
p9 <- ggplot(df, aes(x=studytime, y=final.grade, fill=factor(studytime))) + geom_boxplot()
p9
```

```{r}
head(df)
```

```{r}
library(dplyr)

study.df <- df %>%
  mutate(studytime = case_when(
    studytime == 1 ~ "<2",
    studytime == 2 ~ "2-5",
    studytime == 3 ~ "5-10",
    studytime == 4 ~ ">10",
    TRUE ~ as.character(studytime)
  ))
```


```{r}
head(study.df)
```

```{r}
p10colors <- viridisLite::viridis(4)

p10colors
show_col(p10colors)

p10colors_r <- c("#FDE725FF", "#35B779FF", "#31688EFF", "#440154FF")

```


```{r}
library(plotly)

p10 <- plot_ly(study.df, x = ~absences, y = ~final.grade, color = ~studytime, colors = p10colors_r,
               text = ~paste(studytime, "hours spent studying per week"), 
               hoverinfo = "text", type="scatter", mode="markers", marker = list(opacity = 0.7, size=8)) %>% 
  layout(title = "Total Absences vs. Final Grade", 
         xaxis = list(title = "Absences"), 
         yaxis = list(title = "Final Grade"), 
         showlegend = FALSE)

p10
```


```{r}
ggplot(df, aes(x = final.grade, y = factor(weekday.alc), fill = factor(weekday.alc))) + 
  geom_density_ridges() + 
  scale_fill_viridis_d(5) +
  theme_minimal() + 
  theme(legend.position = "none")
  scale_y_discrete(breaks = 1:5, labels = c("very low", "low", "moderate", "high", "very high")) 
```

```{r}
dim(df)
```

# Plot 11
```{r}
library(tidyr)
```



### Plot 11b

```{r}
grades_df <- df %>% select(first.period.grade, second.period.grade, final.grade, health.status)
grades_df <- gather(grades_df, period, grade, 1:3)
grades_df$Period <- grades_df$period
grades_df$Period <- factor(grades_df$Period, levels = c("first.period.grade", "second.period.grade", "final.grade"))

grades_df <- grades_df %>%
  mutate(Period = case_when(
    Period == "first.period.grade" ~ "1st Semester",
    Period == "second.period.grade" ~ "2nd Semester",
    Period == "final.grade" ~ "Final",
    TRUE ~ as.character(Period)
  ))

grades_df <- grades_df %>%
  mutate(health.status = case_when(
    health.status == 1 ~ "Very Good",
    health.status == 2 ~ "Good",
    health.status == 3 ~ "Moderate",
    health.status == 4 ~ "Poor",
    health.status == 5 ~ "Very Poor",
    TRUE ~ as.character(health.status)
  ))

grades_df$health.status <- factor(grades_df$health.status, levels = c("Very Good", "Good", "Moderate", "Poor", "Very Poor"))

head(grades_df)
```


```{r}
p11colors <- c("#2D708EFF", "#2D708EFF", "grey", "grey", "grey")
```


```{r}
p11 <- plot_ly(grades_df, x = ~factor(health.status), y = ~grade, type = "violin", frame = ~Period,
             color = ~factor(health.status), colors = p11colors) %>% 
  layout(title = list(text = "Grade Distribution by Student's Health Condition over Semester", size = 20),
            xaxis = list(title = "Health Status", titlefont = list(size = 16), tickfont = list(size = 14)),
            yaxis = list(title = "Grade", range = c(0, 100), titlefont = list(size = 16), tickfont = list(size = 14), dtick = 10),
            boxmode = "group", 
            showlegend = FALSE) %>%
  animation_opts(frame = 1000,  # Controls speed of transition; higher number = slower
                 redraw = TRUE, # Redraws the plot in each frame
                 easing = "linear") %>% # FIX THIS PART
  
  layout(
   updatemenus = list( 
     list(
       type = "buttons", 
       showactive = FALSE,
       x = 0, 
       y = 1, 
       buttons = list(
         
        list(
           label = "Pause", # Name of the label
           method = "animate", # What do to when the button is clicked
           args = list(NULL, list(frame = list(duration = 0,  # Stops immediately
                                               redraw = TRUE), 
                                  mode = "immediate"))
         )
       )
     )
   )
 ) 
    

p11
```

# Plot 12
```{r}
p12colors <- viridisLite::viridis(12)
show_col(p12colors)
p12colors

# p12colors <- c("#C2DF23FF", "#482173FF")
```
```{r}
edu_df <- df %>%
  mutate(higher.edu = case_when(
    higher.edu == 0 ~ "No", 
    higher.edu == 1 ~ "Yes", 
    TRUE ~ as.character(higher.edu)
  ))

edu_df <- edu_df %>%
  mutate(mom.edu = case_when(
    mom.edu == "no education" ~ "No education", 
    mom.edu == "elementary school" ~ "Elementary school",
    mom.edu == "middle school" ~ "Middle school",
    mom.edu == "high school" ~ "High school", 
    mom.edu == "university or higher" ~ "University or higher", 
    TRUE ~ as.character(mom.edu)
  ))

edu_df <- edu_df %>%
  mutate(dad.edu = case_when(
    dad.edu == "no education" ~ "No education", 
    dad.edu == "elementary school" ~ "Elementary school",
    dad.edu == "middle school" ~ "Middle school",
    dad.edu == "high school" ~ "High school", 
    dad.edu == "university or higher" ~ "University or higher", 
    TRUE ~ as.character(dad.edu)
  ))

edu_df$mom.edu <- factor(edu_df$mom.edu, levels = c("No education", "Elementary school", "Middle school", "High school", "University or higher"))

edu_df$dad.edu <- factor(edu_df$dad.edu, levels = c("No education", "Elementary school", "Middle school", "High school", "University or higher"))

# edu_df$higher.edu <- factor(edu_df$higher.edu, levels = c("Yes", "No"))

head(edu_df)
```

```{r}
table(df$mom.edu)
```


```{r}
ggplot(edu_df, aes(y=mom.edu, fill=factor(higher.edu))) + geom_bar() + theme_minimal() + labs(title = " Students Interested in Higher Education by Mother's Education Level", x = "Count", y = "Mother's Highest Education Level") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_manual(values = c("grey", "#2D708EFF"))
```

# Plot 13

```{r}
strip_df <- df
strip_df$activities <- as.factor(strip_df$activities)
```


```{r}
stripchart(final.grade~activities, 
           data = strip_df, 
           main = "Final Grade vs. Activities", 
           # xlab = "Final Grade", 
           # ylab = "Activities", 
           method = "jitter", 
           col = c("yellow", "blue"),
           group.names = c("Yes", "No"),
           pch = 1) # default 
```


# Plot 14 
```{r}
ggplot(df, aes(x=factor(freetime), y=final.grade)) + 
  geom_boxplot(aes(fill=factor(freetime))) + 
  theme_minimal() + 
  scale_fill_viridis_d() +
  facet_grid(.~activities, labeller = as_labeller(c("0" = "No Extracurriculars", "1" = "Extracurriculars"))) + 
  guides(fill = FALSE) + 
  labs(title = "Final Grade Distribution by How Freetime is Spent", x = "Amount of Freetime after School", y = "Final Grade") + 
  scale_x_discrete(breaks = 1:5, labels = c("Very less", "Less", "Some", "More", "A lot more")) + 
  theme(axis.text.x = element_text(angle = 35), 
        plot.title = element_text(hjust = 0.5, size = 18), 
        axis.title = element_text(size = 14),
        strip.text = element_text(size = 14), 
        axis.text = element_text(size = 10))
```

# Plot 15

```{r}
library(ggalluvial)
```

```{r}
alluv_df <- df %>% 
  group_by(romantic, famrel, health.status) %>% 
  summarize(freq = n())

alluv_df
```

```{r}
ggplot(alluv_df, 
       aes(y = freq, axis1 = factor(romantic), axis2 = factor(famrel), axis3 = factor(health.status))) + geom_alluvium(aes(fill = factor(health.status)), width = 0, knot.pos = 1/4, reverse = FALSE) + 
  guides(fill = "none") +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", aes(label= after_stat(stratum)), reverse = FALSE) +
  scale_x_continuous(breaks = 1:3, labels = c("Romantic Partner", "Family Relationship Status", "Health Condition"))

```

```{r}
alluv_df$romantic <- factor(alluv_df$romantic, labels = c("No", "Yes"))
alluv_df$famrel <- factor(alluv_df$famrel, labels = c("Very poor", "Poor", "Decent", "Good", "Very Good"))
```


```{r}
ggplot(alluv_df, 
       aes(y = freq, axis2 = factor(romantic), axis1 = factor(famrel))) + 
         geom_alluvium(aes(fill = factor(famrel))) + 
  guides(fill = "none") +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", aes(label= after_stat(stratum)), reverse = FALSE) +
  scale_x_continuous(breaks = 1:2, labels = c("Quality of Family Relationships", "Romantic Partner")) + 
  labs(title = "Satisfaction with Family Relationships by Romantic Relationships", y = "Frequency") + 
  scale_fill_viridis_d() +
  theme_minimal() + 
  theme(plot.title = element_text(size = 18, hjust = 0.5), 
        axis.title.y = element_text(size = 16), 
        axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 12))
```

# Plot 16 

```{r}
F_df <- df[df$sex == "F", ] 
F_cor <- cor(F_df$drinking, F_df$age)

M_df <- df[df$sex == "M", ] 
M_cor <- cor(M_df$drinking, M_df$age)
```

```{r}
corr <- bquote(R^2)
print(corr)
```


```{r}
p16_F <- ggplot(F_df, aes(x = age, y = drinking, color = drinking)) + 
  geom_point(position = "jitter") +
  geom_smooth(method = "lm", color = "darkgrey") +
  scale_color_viridis() + 
  theme_minimal() + 
  geom_text(aes(x = max(F_df$age), y = max(F_df$drinking), label = paste("Corr:", round(F_cor, 3))), 
            hjust = 1, vjust = 1.5, size = 4, color = "black") 
```
```{r}
p16 <- ggplot(M_df, aes(x = age, y = drinking, color = drinking)) + 
  geom_point(position = "jitter") +
  geom_smooth(method = "lm", color = "darkgrey") +
  scale_color_viridis() + 
  theme_minimal() + 
  geom_text(aes(x = max(M_df$age), y = max(M_df$drinking), label = paste("Corr:", round(M_cor), 3)), hjust = 1, vjust = 1.5, size = 4, color = "black")

p16
```
```{r}
library(ggpubr)
```

```{r}
ggplot(df, aes(x = age, y = drinking, color = drinking)) +
  geom_point(position = "jitter") + 
  scale_color_viridis() + 
  theme_minimal() +
  geom_smooth(method = "lm", color = "black") +
  stat_cor(method = "pearson", r.accuracy = 0.01, p.accuracy = 0.001, label.x.npc = 0.5) + 
  facet_grid(.~sex)
```


# What are some factors associated with students' health and academic success? 

### Plot 1: Relationships between Numeric Variables 

Plot 1 is a correlogram containing the correlation coefficients between 
all the numeric variables in the data. The plot serves as a high-level starting point 
for understanding the data. Aside from obvious cases of high correlation 
between variables where multicollinearity is present (such as weekend alcohol 
consumption and overall drinking habits), the correlogram reveals relationships 
between several variables that may be interesting to explore. \newline 
Some of the relationships of interest were final grade and alcohol consumption ($r = -0.13$), 
final grade and total absences ($r = -0.05$), and hours spent studying per week 
and final grade ($r = 0.16$). The correlogram provides a good starting point for 
examining which factors influence high school students' health and academic success,
but it is limiting for visualizing relationships between variables because it 
doesn't capture categorical variables or outliers.


```{r echo = FALSE}
library(ggcorrplot)

num_vars <- df %>% select(age, traveltime, studytime, failures, famrel, 
                          freetime, goout, weekday.alc, weekend.alc, 
                          health.status, absences, final.grade, drinking)

cor_matrix <- cor(num_vars)
cor_df <- as.data.frame(cor_matrix)

p1 <- ggcorrplot(cor_df, type = "lower", lab = TRUE, lab_size = 3) + 
  theme_minimal() + 
  # theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  labs(x = "", y = "", title = "Correlation between Numeric Variables", fill = "Correlation") + 
  theme(plot.title = element_text(size = 20, hjust = 0.5), 
        axis.text = element_text(size = 14), 
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

p1_color_scale <- scale_fill_gradient2(low = "#FCD225FF", mid = "white", high = "#0D0887FF", midpoint = 0, limits = c(-1, 1), na.value = "white")

p1 <- p1 + p1_color_scale

p1
```





